Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2015-08-13 article.Rmd

This article describes a new method for inferring the gender of personal names using large historical datasets. In contrast to existing methods of predicting gender that treat names as if they are timelessly associated with gender, this method uses a historical approach that takes into account the dynamic nature of naming practices. It uses historical data to measure the likelihood that a name was associated with a particular gender based on the time or place under study. This approach generates more accurate results for sources that encompass changing periods of time, providing digital humanities scholars with a tool to estimate the gender of names. The article first describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). It goes on to apply the method to a case study in which we examine gender and gatekeeping in the American historical profession over the past half-century. The Gender package illustrates the importance of incorporating historical approaches to computer science and related fields.

@@

As humanities datasets grow larger and larger, researchers have not only been able to access information that is there, but to infer information that isn't there. Topic modeling, for example, has been used to uncover scholarship trends over a century of academic articles in literary studies. [Goldstone and Underwood 2014]. Can researchers similarly use personal names to infer information about individual people? After all, personal names appear across a wide range of sources in the humanities: literary characters in stories and novels, authorship information in archival metadata, or individuals mentioned by historical newspapers and magazine articles. Authors, characters, or historical actors frequently stand at the analytical center of humanities scholarship, and their names provide an intriguing way to infer additional information about them.^[The availability of large-scale datasets dealing with personal information raises important ethical issues for digital humanists, particularly in light of the revelations about domestic spying at the NSA brought forth by Edward Snowden in 2013. [Weingart 2014] Kieran Healy has used network analysis to demonstrate how researchers can infer information from datasets while also touching on the troubling ethical issue of surveillance [Healy 2013].]

One of the central kinds of information researchers can infer from a personal name is the gender of that individual: "Jane Fay," for instance, is almost certainly a woman, while her brother "John Fay" is almost certainly a man. A computer program can infer their respective genders by looking up “Jane” and “John” in an existing dataset that matches names to genders. But this approach runs into a problem: the link between gender and naming practices, like language itself, is not static. What about Jane and John Fay's sibling, Leslie? If Leslie was born in the past sixty years, chances are good that Leslie would be their sister. But if the three siblings were born in the early twentieth century, chances are good that Leslie would be their brother. That is because the conventional gender for the name Leslie switched over the course of one hundred years. In 1900 some 92% of the babies born in the United States who were named Leslie were male, while in 2000 about 96% of the Leslies born in that year were female.

@@

For those working on contemporary subjects the "Leslie problem" is not an especially pressing one. A film scholar analyzing male and female characters in modern American television shows, for instance, can turn to a variety of tools that use current databases of names, and these tools will likely identify the gender of Leslie Fay correctly. One online service, Genderize.io, determines the gender of first names "from user profiles across major social networks." [Genderize.io n.d.] Checking the service for the name Leslie returns a prediction that the name is female, along with an estimate of male/female proportions of the uses among members of unnamed social media sites.^[In this example the Genderize.io service has been accessed through an API wrapper provided in our Gender package described below.]

@@

But researchers studying a longer timespan need to take into account the changing nature of naming practices. If that same film scholar wanted to compare the characters of modern television shows to characters during the silent film era,  existing tools like Genderize.io would misidentify Leslie Fay's gender in the earlier screenplays. Other names used in the United States such as Madison, Morgan, Sydney, and Kendall, have also flipped their gender over the past century and would similarly be misidentified. The prevalence of changes in the gender of names in the last half of the twentieth century implies that gendered lists of names drawn from contemporary data are anachronistic and so of limited use for datasets that encompass changing periods of time.^[The few exceptions of names which flip from being female to male, such as "Jan", have a different causal mechanism. We speculate that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommon, other causes besides changing gender preferences can explain the change.]

@@

The "Leslie problem" is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match the groups used by a contemporary tool such as Genderize.io will also encounter this problem. For example, a literary scholar studying a corpus of poems from the mid-twentieth century might wish to compare the stanza structures of male and female poets. Using a tool like Genderize.io based on contemporary data risks misidentifying these writers, many of whom were born (and named) more than a century prior to the creation of these modern name datasets. This problem will only increase with growing life expectancies: as of 2012, the average American lifespan was nearly seventy-nine years---more than enough time for naming practices to change quite dramatically between a person’s birth and death [World Bank 2014]. Predicting gender from first names therefore requires a dynamic method that takes into change over time.

Our solution to the "Leslie problem" is to create a software package that combines a predictive algorithm with several historical datasets suitable to various times and regions. The algorithm calculates the proportion of male or female uses of a name in a given birth year or range of years. It thus can provide not only a prediction of the gender of a name, but also a measure of that prediction's likelihood. For example, our program predicts that a person named Leslie born in 1950 is female, but only just barely: 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. Researchers can use these probabilities to determine what level of certainty is acceptable for predicting the gender of different names.

@@

As with any tool, researchers who use this method need to be fully aware of the ways it replicates a problematic, state-defined gender binary. The onus is on the user to understand how its underlying data was gathered, what it implies, and how it reflects social and cultural practices. Imperfect as it may be, this method nevertheless gives digital humanities scholars an important means of studying gender in textual collections. We have made the method as transparent and flexible as possible through the inclusion of probability figures - ex. 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. This makes the underlying data visible to users and allows them the opportunity to interrogate the assumptions that come with it. When used thoughtfully, this method can infer additional information about individuals who stand at the heart of humanities scholarship, providing another way to see and interpret sources across large scales of analysis.

@@

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the painful limitations of this sea change for female historians: ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck's report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession [Lunbeck 2005]. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period. We study one of the bedrocks of the historical profession, scholarly research, and focus on two types of scholarly output: the dissertation and the monograph. Although women have achieved rough parity with men in terms of the number of completed dissertations, female historians continue to face an unequal playing field in the monographic landscape. Applying our method to the pages of the *American Historical Review* reveals a persistent and significant gap in the number of male-authored and female-authored books reviewed by one of the field’s premier academic journals.

@@

The digital turn has made available very large datasets, yet in the midst of abundance many humanists continue to deal with the problem of incompleteness and scarcity within their sources. Our method uses *abundant* existing datasets to infer additional information from *scarce* datasets. Its major contribution is to incorporate a historical approach to this algorithm, one that takes into account how gendered naming practices have changed over time. This temporal methodology provides much improved results over simpler, anachronistic, or ahistorical lookup methods. The Gender package is, of course, far from perfect. By relying on historical datasets collected by state agencies, it reinforces a gender binary and larger power structure that obscures how people may identify themselves and how they experience the world. To borrow a phrase from Miriam Posner, the Gender package does not meet the "radical, unrealized potential of digital humanities." But it might.

The Gender package is entirely open-source, and we have provided detailed and transarent documentation of its code and data. Researchers who wish to build upon it can modify the package accordingly. To take one imagined scenario: Facebook allows its users to either select from a large spectrum of gender identities or to write in their own prefered identity. Because these identities are associated with the user's first name and date of birth, this is exactly the kind of historical dataset that could be incorporated into the Gender package. Doing so would replace the package's simplistic categories of "Male" and "Female" with ones like "Female to Male", "Bigender", or "Cisgender Female". Although this imagined scenario would create its own set of challenges, it would represent a major improvement in how the Gender package dealt with the complexity and fluidity of gender identity. ^[This imagined scenario runs into its own set of problems. Facebook's business model makes it unlikely to grant researchers open access to its proprietary and closely-guarded datasets. And even if it did grant access to this data, Facebook collected it for particular purposes and in particular ways. Just as government agencies like the Social Security Administration collect data to serve the purposes of the American state, Facebook's data collection reflects the priorities of a global corporation rather than the lived experience of its users.] By building as much flexibility and transparency into the method as possible, we hope that others can continue to make just these kinds of improvements.

Even as it currently stands, the Gender package meets a pressing methodological need. Since its initial release in February 2014, the Gender package has been downloaded more than ________ times [??????] and incorporated as a pedagogical tool into digital humanities courses [Jockers 2014]. Researchers have used it to uncover gender inequity in a range of different areas, from biases in students' teaching evaluations to disparities between attendees and authors at the annual Digital Humanities Conference [Schmidt 2015, Weingart 2015]. These implementations help illustrate the value of the method. Its major intervention is to algorithmically inferring gender in textual sources in a way that takes into account change over time. Rather than trying to shoehorn existing computational methods, the Gender package uses a humanistic starting point---the historical relationship between gender and naming practices---to contribute a new method to computer science. This reverses the traditional disciplinary direction between the humanities and computer science.

@@

[Argamon et. al. 2009] Argamon, Shlomo, Jean-Baptiste Goulain, Russell Horton, and Mark Olsen. "Vive la Différence! Text Mining Gender Difference in French Literature" *Digital Humanities Quarterly*. 3.2 (2009). <http://www.digitalhumanities.org/dhq/vol/3/2/000042/000042.html>

@@

[Orlando Project n.d.] Orlando Project. "About." <http://www.artsrn.ualberta.ca/orlando/?page_id=2>. Accessed 23 June 2015.]

@@

[R Core Team 2014-] R Core Team. *R: A language and environment for statistical computing*. R Foundation for Statistical Computing, Vienna, Austria, 2014. <http://www.R-project.org/>

@@

[Townsend 2008] Townsend, Robert B. "The Status of Women and Minorities in the History Profession, 2008." *Perspectives on History*. September 2008, <http://www.historians.org/publications-and-directories/perspectives-on-history/september-2008/the-status-of-women-and-minorities-in-the-history-profession-2008>

[Townsend 2012] Townsend, Robert B. "What Makes a Successful Academic Career in History? A Field Report from the Higher Ranks." *Perspectives on History*. December 2012, <http://www.historians.org/publications-and-directories/perspectives-on-history/december-2012/what-makes-a-successful-academic-career-in-history>

@@

[World Bank 2014-] World Bank. "Life expectancy at birth, total (years)." *World Bank Open Data*. <http://data.worldbank.org/indicator/SP.DYN.LE00.IN>. Accessed 12 September 2014

>>>>

This article describes a new method for inferring the gender of personal names using large historical datasets. In contrast to existing methods of predicting gender that treat names as if they are timelessly associated with gender, this method uses a historical approach that takes into account the dynamic nature of naming practices. It uses historical data to measure the likelihood that a name was associated with a particular gender based on the time or place under study. This approach generates more accurate results for sources that encompass changing periods of time, providing digital humanities scholars with a tool to estimate the gender of names. The article first describes the methodology as implemented in the [Gender](https://github.com/ropensci/gender) package for the [R programming language](http://www.r-project.org/). It goes on to apply the method to a case study in which we examine gender and gatekeeping in the American historical profession over the past half-century. The Gender package illustrates the importance of incorporating historical approaches into computer science and related fields.

@@

As humanities datasets grow larger and larger, researchers have not only been able to access information that *is* there, but to infer information that *isn't* there. Topic modeling, for example, has been used to uncover underlying trends across a century of academic articles in literary studies. [Goldstone and Underwood 2014]. Can researchers similarly use personal names to infer information about individual people? After all, personal names appear across a wide range of sources in the humanities: literary characters in stories and novels, authorship information in archival metadata, individuals mentioned by historical newspapers and magazine articles. Authors, characters, or historical actors are the kind of individuals who stand at the analytical center of humanities scholarship, and their names provide an intriguing way to infer additional information about them.^[The availability of large-scale datasets dealing with personal information raises important ethical issues for digital humanists, particularly in light of the revelations about domestic spying at the NSA brought forth by Edward Snowden in 2013. [Weingart 2014] Kieran Healy has used network analysis to demonstrate how researchers can infer information from datasets while also touching on the troubling ethical issue of surveillance [Healy 2013].]

One of the basic pieces of information researchers can infer from a personal name is the gender of that individual: "Jane Fay," for instance, is likely a woman, while her brother "John Fay" is likely a man. A computer program can infer their respective genders by looking up “Jane” and “John” in an existing dataset that matches names to genders. But this approach runs into a problem: the link between gender and naming practices, like language itself, is not static. What about Jane and John Fay's sibling, Leslie? If Leslie was born in the past sixty years, chances are good that Leslie would be their sister. But if the three siblings were born in the early twentieth century, chances are good that Leslie would be their brother. That is because the conventional gender for the name Leslie switched over the course of one hundred years. In 1900 some 92% of the babies born in the United States who were named Leslie were classified as male, while in 2000 about 96% of the Leslies born in that year were classified as female.

@@

For those working on contemporary subjects, the "Leslie problem" is not an especially pressing one. There are a variety of tools that use current databases of names, including Genderize.io, which predicts the gender of a first name from the user profiles of unnamed social media sites [Genderize.io n.d.]. A film scholar analyzing male and female characters in a database of modern American television shows might use Genderize.io to accurately predict the gender of these characters, including Leslie Fay. **do we need this footnote?^[In this example the Genderize.io service has been accessed through an API wrapper provided in our Gender package described below.]** But researchers studying a longer timespan need to take into account the changing nature of naming practices. If that same film scholar wanted to compare the characters of modern television shows to characters during the silent film era,  existing tools like Genderize.io would misidentify Leslie Fay's gender in the earlier screenplays. Other names used in the United States such as Madison, Morgan, Sydney, and Kendall, have also flipped their gender over the past century and would similarly be misidentified. The prevalence of changes in the gender of names in the last half of the twentieth century implies that gendered lists of names drawn from contemporary data are anachronistic and so of limited use for datasets that encompass changing periods of time.^[Most names flip from male to female. The few exceptions, such as "Jan", have a different causal mechanism. We speculate that Jan as an English name is female but that Jan as a Scandinavian name is male. Because the name has remained relatively uncommon, other causes besides changing gender preferences can explain the change.]

@@

@@

The "Leslie problem" is not just for researchers who think of themselves as historians. Anyone studying a period longer than a few years, or anyone studying a group whose demographics do not match the groups used by a contemporary tool such as Genderize.io will also encounter this problem. A literary scholar studying a corpus of poems from the mid-twentieth century might wish to compare the stanza structures of male and female poets. Existing tools based on contemporary data risks misidentifying these writers, many of whom were born (and named) more than a century prior to the creation of these modern name datasets. This problem will only increase with growing life expectancies: as of 2012, the average American lifespan was nearly seventy-nine years---more than enough time for naming practices to change quite dramatically between a person’s birth and death [World Bank 2014]. Predicting gender from first names therefore requires a dynamic method that takes into change over time.

Our solution to the "Leslie problem" is to create a software package that combines a predictive algorithm with several historical datasets suitable to various times and regions. The algorithm calculates the proportion of male or female uses of a name in a given birth year or range of years. It thus can provide not only a prediction of the gender of a name, but also a measure of that prediction's likelihood. For example, our program predicts that a person named Leslie who was born in 1950 is female, but only just barely: 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. Researchers can use these probabilities to determine what level of certainty is acceptable for predicting the gender of different names.

@@

Researchers who use this method need to be fully aware of the ways it replicates a problematic, state-defined gender binary. The onus is on the user to understand how its underlying data was gathered, what it implies, and how it reflects social and cultural practices. Imperfect as it may be, this method nevertheless gives digital humanities scholars an important means of studying gender in textual collections. We have made the method as transparent and flexible as possible through the inclusion of probability figures - ex. 52 percent of the babies named Leslie born in the United States in 1950 were girls and 48 percent were boys. This makes the underlying data visible to users and allows them the opportunity to interrogate the assumptions that come with it. When used thoughtfully, this method can infer additional information about individuals who stand at the heart of humanities scholarship, providing another way to see and interpret sources across large scales of analysis.

@@

In a 2005 report for the American Historical Association, Elizabeth Lunbeck acknowledged "a sea change in the [historical] profession with respect to gender" before going on to describe the limits to this change for female historians, who face ongoing personal discrimination, lower salaries, and barriers to securing high-ranking positions. Lunbeck's report drew in part on a survey of 362 female historians that produced a rich source of responses detailing the deep and multi-faceted challenges facing women in the profession [Lunbeck 2005]. What follows is a quantitative supplement to Lunbeck's analysis that uses our program to analyze gender representation amongst historians across a much larger scale and a much longer time period. We study one of the bedrocks of the historical profession --- scholarly research --- and focus on two types of output: the dissertation and the monograph. Although women have achieved rough parity with men in terms of the number of completed dissertations, female historians continue to face an unequal playing field in the monographic landscape. Applying our method to the pages of the *American Historical Review* reveals a persistent and significant gap in the number of male-authored and female-authored books reviewed by one of the field’s premier academic journals.

@@

The digital turn has made large datasets widely available, yet in the midst of abundance many humanists continue to deal with the problem of incompleteness and scarcity within their sources. Our method uses *abundant* existing datasets to infer additional information from *scarce* datasets. Its major contribution is to incorporate a historical method, one that takes into account how gendered naming practices have changed over time. This temporal approach provides much improved results over simpler, anachronistic, or ahistorical lookup methods. The Gender package is, of course, far from perfect. By relying on historical datasets collected by state agencies, it reinforces a gender binary and larger power structure that obscures how people may identify themselves and how they experience the world. To borrow a phrase from Miriam Posner, the Gender package does not necessarily meet the "radical, unrealized potential of digital humanities" [Posner 2015]. But it might.

The Gender package is entirely open-source, and we have provided detailed documentation for its code and underlying data so that researchers who wish to build upon it can modify the package accordingly. To take one imagined scenario: Facebook allows its users to select from a spectrum of gender identities or to write in their own prefered identity. Because these identities are associated with the user's first name and date of birth, this is exactly the kind of historical dataset that could be incorporated into the Gender package. Doing so would replace the package's simplistic categories of "Male" and "Female" with more detailed categories like "Female to Male", "Bigender", or "Cisgender Female". Although this imagined scenario would create its own set of challenges, it would represent a major improvement in how the Gender package dealt with the complexity and fluidity of gender identity.^[This imagined scenario runs into its own set of problems. Facebook's business model makes it unlikely to grant researchers open access to its proprietary and closely-guarded datasets. And even if it did grant access to this data, Facebook collected it for particular purposes and in particular ways. Just as government agencies like the Social Security Administration collect data to serve the purposes of the American state, Facebook's data collection more accurately reflects the priorities of a global corporation rather than the lived experience of its users.] By building as much flexibility and transparency into the method as possible, we hope that others can continue to make just these kinds of improvements.

As it currently stands, the Gender package meets a pressing methodological need. Since its initial release in February 2014, the Gender package has been downloaded more than **________ times [??????]** and incorporated as a pedagogical tool within digital humanities courses [Jockers 2014]. Researchers have used it to uncover instances gender inequity in a range of different areas, from biases in students' teaching evaluations to disparities between attendees and authors at the annual Digital Humanities Conference [Schmidt 2015, Weingart 2015]. These implementations help illustrate the value of the method. Its major intervention is to algorithmically inferring gender in textual sources in a way that takes into account change over time. Rather than trying to shoehorn existing computer science methods to fit humanities research, the Gender package reverses this traditional disciplinary relationship. It uses a humanistic starting point---the historical relationship between gender and naming practices---to contribute a new method to computer science.

@@

[Argamon et. al. 2009] Argamon, Shlomo, Jean-Baptiste Goulain, Russell Horton, and Mark Olsen. "Vive la Différence! Text Mining Gender Difference in French Literature" *Digital Humanities Quarterly*. 3.2 (2009). <http://www.digitalhumanities.org/dhq/vol/3/2/000042/000042.html>.

@@

[Orlando Project n.d.] Orlando Project. "About." <http://www.artsrn.ualberta.ca/orlando/?page_id=2>. Accessed 23 June 2015.

@@

[R Core Team 2015-] R Core Team. *R: A language and environment for statistical computing*. R Foundation for Statistical Computing, Vienna, Austria, 2015. <http://www.R-project.org/>.

@@

[Townsend 2008] Townsend, Robert B. "The Status of Women and Minorities in the History Profession, 2008." *Perspectives on History*. September 2008, <http://www.historians.org/publications-and-directories/perspectives-on-history/september-2008/the-status-of-women-and-minorities-in-the-history-profession-2008>.

[Townsend 2012] Townsend, Robert B. "What Makes a Successful Academic Career in History? A Field Report from the Higher Ranks." *Perspectives on History*. December 2012, <http://www.historians.org/publications-and-directories/perspectives-on-history/december-2012/what-makes-a-successful-academic-career-in-history>.

@@

[World Bank 2014-] World Bank. "Life expectancy at birth, total (years)." *World Bank Open Data*. <http://data.worldbank.org/indicator/SP.DYN.LE00.IN>. Accessed 12 September 2014.
